---
# PA1_Template (Reproducible Research, Week 2)

### Author: "SB"
### date: "Tuesday, May 12, 2015"

The first step is to download the .zip file locally into the working directory and load it into a data frame.

- Install 'downloader' package

- download .zip file locally, and unzip it

- read.csv into a dataframe

```{r download data, echo=FALSE, warning=FALSE, message=FALSE}

       install.packages("downloader", 
                        repos = "http://cran.us.r-project.org",
                        silent = TRUE)
       install.packages("dplyr", 
                        repos = "http://cran.us.r-project.org",
                        silent = TRUE)
       install.packages("plyr", 
                        repos = "http://cran.us.r-project.org",
                        silent = TRUE)
       install.packages("lubridate", 
                        repos = "http://cran.us.r-project.org",
                        silent = TRUE)
       
       library(downloader, quietly=TRUE)
       require(dplyr, quietly=TRUE)
       library(plyr, quietly=TRUE)
       library(lubridate, quietly=TRUE)

       #Load file location              
       url <- "https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2Factivity.zip"
       
       download(url, dest="repdata-data-activity.zip", mode="wb")
       
       unzip("repdata-data-activity.zip", exdir = "./repdata-data-activity")
       
       
       #Load file location       
       file <- "./repdata-data-activity/activity.csv"
       
       
       #Read file into a data frame       
       activityFile <- read.csv(file, header=TRUE)
       activity <- activityFile       

       #Remove 'NA' from steps
       activity <- subset(activity, !is.na(activity$steps))
       
       #Convert date into a date data type
       activity$date <- as.Date(as.character(activity$date))



```

____________________________________________________________________________________________________

### Let us now sum the steps by each day and create a histogram

```{r process Data part1, echo=FALSE}
       
       
       sumActivity <- aggregate(activity$steps, by=list(activity$date), FUN=sum)
       names(sumActivity) <- c("date", "steps")     

       
       hist(x=sumActivity$steps, 
            col="gray", 
            main = "Steps taken Daily", 
            xlab = "Steps taken daily")    
     

```

### Calculate the Mean and Median steps taken per day

```{r mean&median, echo=FALSE}

       mean <- mean(sumActivity$steps, na.rm=TRUE)
       median <- median(sumActivity$steps, na.rm=TRUE)

```
Mean: `r mean`

Median: `r median`

_____________________________________________________________________________________________________

### Make a plot for Average Daily Pattern of steps taken each day

```{r averageDaily Pattern, echo=FALSE}


       stepsByInterval <- aggregate(activity$steps, by=list(activity$interval), FUN=mean)
       names(stepsByInterval) <- c("interval", "steps")
       

       plot(x = stepsByInterval$interval,
            y = stepsByInterval$steps,
            type = "l",
            main = "Daily Avg steps by Interval",
            xlab = "Interval",
            ylab = "Daily Avg steps")

       maxInterval <- with(stepsByInterval,interval[which.max(steps)])

```


Highest steps are in `r maxInterval` Interval

_____________________________________________________________________________________________________

### Take a stock of missing values in the data set

```{r missingValues, echo=FALSE}


       missingCount <- sum(is.na(activityFile$steps))
   

```

There are a total of `r missingCount` elements that have missing values in the data set


Lets consider filling up the missing values with the mean for that interval

```{r fillMissingValues, echo=FALSE}

       activityUpdated <- activityFile
     

       for (n in 1:nrow(activityUpdated)){
              
              if (is.na(activityUpdated$steps[n])) {
                     
                row_num <- which(stepsByInterval$interval == activityUpdated$interval[n])     
                activityUpdated$steps[n] <- stepsByInterval$steps[row_num]
                                             
              }
              
       }
      
      

       sumActivity <- aggregate(activityUpdated$steps, by=list(activityUpdated$date), FUN=sum)
       names(sumActivity) <- c("date", "steps")     

       
       hist(x=sumActivity$steps, 
            col="gray", 
            main = "Steps taken Daily", 
            xlab = "Steps taken daily")    
     
       mean1 <- mean(sumActivity$steps, na.rm=TRUE)
       median1 <- median(sumActivity$steps, na.rm=TRUE)

       diffMean <- mean1-mean
       diffMedian <- median1-median

```

Mean Old: `r mean` 

Mean New: `r mean1`

Shift in Mean: `r diffMean`

Median Old: `r median`

Median New: `r median1`

Shift in Median: `r diffMedian`
___________________________________________________________________________________________________


```{r weekdays, echo=FALSE}
       

       activityUpdated$weekday <- weekdays(as.Date(activityUpdated$date))
       
       
       for (i in 1:nrow(activityUpdated)) {
              
              if (activityUpdated$weekday[i] == "Saturday" | activityUpdated$weekday[i] == "Sunday") {
                     activityUpdated$weekday[i] <- "weekend"                     
              }
              else{
                    activityUpdated$weekday[i] <- "weekday"    
              }
                     
       }
       
       activityUpdated$weekday <- factor(activityUpdated$weekday)
       
       activityWeekday <- subset(activityUpdated, activityUpdated$weekday=="weekday")
       activityWeekday <- aggregate(activityWeekday$steps, by=list(activityWeekday$interval), FUN=mean)
       names(activityWeekday) <- c("interval", "steps")

       
       activityWeekend <- subset(activityUpdated, activityUpdated$weekday=="weekend")
       activityWeekend <- aggregate(activityWeekend$steps, by=list(activityWeekend$interval), FUN=mean)
       names(activityWeekend) <- c("interval", "steps")
       

       par(mfrow = c(2, 1))
       
       plot(x = activityWeekday$interval, 
            y = activityWeekday$steps,
            type="l", 
            xlab="5 minute interval", 
            ylab="Weekday steps", 
            main="Weekdays")

      
       plot(x = activityWeekend$interval, 
            y = activityWeekend$steps,
            type="l", 
            xlab="5 minute interval", 
            ylab="Weekend steps", 
            main="Weekends")

```

As per data seems like there is more physical activity during the weekends, as compared to weekdays.